Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-07-06 p5/stability_rnn.tex
Second revision.

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol. Whenver
a symbol stands for a vector, it denotes a column vector.

@@

We now define a general ODERNN\cite{niu2019recurrent}.
\begin{defn}\label{s7d1}
An ODE-RNN of $n$-th order in non-linearity and $t$-th order in gradient,
called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by
the following update rule.
\begin{eqnarray}
K_1 &=& \sigma_1(W_1Y_l + b_1) \label{s7e1} \\
K_q &=& \gamma_q K_{q-1} + \kappa_1\sigma_q\left(
 W_qY_{l-t_{q-1}}+b_q+h\sum_{k=1}^{q-1}\alpha_{q,k}K_k\right)\label{s7e2}\\
Y_{l+1} &=& \gamma_{n+1}Y_l + \kappa_{n+1}\sigma_{q+1}\left(
 W_{n+1}Y_{l-t_{n-1}}+b_{n+1}+h\sum_{k=1}^n\beta_kK_k
 \right)\label{s7e3},
\end{eqnarray}
where
\begin{enumerate}
\item $q = 1, \ldots, n$. This explains the $n$ in the $(n,t)$-ODERNN.
\item The right hand side of equations \eqref{s7e1} to \eqref{s7e3} 
depend on $Y_l$ to $Y_{l - t_{n-1}}$, a total of $n$ previous values.
Here each $t_k = \lfloor tk/n\rfloor$. Because $t_k$ for each hidden layer
is defined in terms of $t$, we also specify it in the name of the ODE. 
This explain $t$ in $(n, t)$-ODERNN.
\item $Y_l \in \mathbb{R}^s$, $l \in \mathbb{Z}$;
\item $W_1,\ldots, W_{n+1}$ are $p \times s$ weight matrices;
\item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors;
\item $\alpha_{q, k}, \beta_k$, for $q = 2, \ldots, n$ and $k = 1, \ldots,
n$, $\gamma_2, \ldots, \gamma_{n}, \kappa_2, \ldots, \kappa_{n}$ are 
$p \times p$ matrices;
\item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
$s \times p$ matrix.
\item $h \in \mathbb{R}$ is a constant and
\item $\sigma_1, \ldots, \sigma_{q+1}$ are activation function defined as
\[
\sigma_i\begin{pmatrix}x_1 \\ \vdots \\ x_p\end{pmatrix} = 
\begin{pmatrix}\sigma_i(x_1) \\ \vdots \\ \sigma_i(x_p)\end{pmatrix},
i = 1, \ldots, q+1.
\]
\end{enumerate}
\end{defn}
\noindent Before proceeding, a few remarks are in order.
\begin{rem}
$K_1, \ldots, K_q$ are all $p$-dimensional vectors. That is, they 
are all members of $\mathbb{R}^p$.
\end{rem}
\begin{rem}
We can readily confirm that the arguments to the activation functions
$\sigma_1, \ldots, \sigma_{q+1}$ are all $p$-dimensional vectors.
\end{rem}
\begin{rem}
The number of hidden layers is $n$. The RNN thus has $n + 2$ layers. 
\end{rem}

@@

\section{Stability analysis}\label{s9}
RNNs frequently have a problem of `exploding gradients'
\cite{gereon2018hands}. In the context of ODEs this is called an 
instability. Because of a correspondence between and RNN and an ODE, we 
can study the problem of exploding gradients in the former as the 
instability in the latter. In this section we will review certain notions
of stability of ODE.

Consider the initial value problem of equation \eqref{s3e1}. The general
$k$-step method to get its approximate numerical solution is defined by
the formula
\begin{equation}\label{s9e1}
\sum_{i=0}^k \alpha_i x_{n+i} = h\sum_{i=0}^k \beta_i f_{n+i},
\end{equation}
where $\alpha_i, \beta_i$ are real constants.
\begin{defn}\label{s9d1}
The $k$-step method defined above is said to be A-stable
\cite{dahlquist1963special} if all solutions of \eqref{s9e1} tend to zero 
as $n \rightarrow \infty$ when the method is applied to the equation
\[
    \frac{dy}{dx} = qx
\]
where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
\eqref{s9d1}.
\end{defn}

In order to introduce the idea of B-stability and BN-stability, we need
the definition of contractive solutions\cite{lambert1991numerical}.
\begin{defn}\label{s9d2}
Let $y_1$ and $y_2$ be any two solutions of \eqref{s3e1} satisfying the
initial conditions $y_1(a) = \eta_1, y_2(a) = \eta_2, \eta_1 \ne \eta_2$.
Then if $||y_1(x_2) - y_2(x_2)|| \le ||y_1(x_1) - y_2(x_1)||$ for all
$a \le x_1 \le x_2 \le b$ then the solutions are said to be contractive
in $[a, b]$.
\end{defn}
A contractive solutions get closer to each other as one gets away from
the initial point.

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because there
is only a single hidden layer. The Burrage-Butcher tensor is thus $-\beta_1
\beta_1^T$, which is \emph{not} positive semidefinite unless $\beta_1 = 0$.
Thus the GRU is not a stable algorithm, an observation also corroborated by
elsewhere\cite{kanai2017preventing}.
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}

>>>>

\section{Notation}\label{s2}
We follow the contemporary mathematical practice of not using separate
notation for vectors and scalars. Instead, we define every symbol. Whenver
a symbol stands for a vector, it denotes a column vector. $\mathbb{R}$ is 
the set of real numbers. $\mathbb{Z}$ is the set of integers. If $z$ is
a complex number then $\Re(z)$ denotes its real part and $\Im(z)$ its
imaginary part.

@@

We now define a general ODERNN\cite{niu2019recurrent}.
\begin{defn}\label{s7d1}
An ODE-RNN of $n$-th order in non-linearity and $t$-th order in gradient,
called $(n,t)$-ODERNN, where $n, t \ge 1$ are integers is described by
the following update rule.
\begin{eqnarray}
K_1 &=& \sigma_1(W_1Y_l + b_1) \label{s7e1} \\
K_q &=& \gamma_q K_{q-1} + \kappa_1\sigma_q\left(
 W_qY_{l-t_{q-1}}+b_q+h\sum_{k=1}^{q-1}\alpha_{q,k}K_k\right)\label{s7e2}\\
Y_{l+1} &=& \gamma_{n+1}Y_l + \kappa_{n+1}\sigma_{q+1}\left(
 W_{n+1}Y_{l-t_{n-1}}+b_{n+1}+h\sum_{k=1}^n\beta_kK_k
 \right)\label{s7e3},
\end{eqnarray}
where
\begin{enumerate}
\item $q = 1, \ldots, n$. This explains the $n$ in the $(n,t)$-ODERNN.
\item The right hand side of equations \eqref{s7e1} to \eqref{s7e3} 
depend on $Y_l$ to $Y_{l - t_{n-1}}$, a total of $n$ previous values.
Here each $t_k = \lfloor tk/n\rfloor$. Because $t_k$ for each hidden layer
is defined in terms of $t$, we also specify it in the name of the ODE. 
This explain $t$ in $(n, t)$-ODERNN.
\item $Y_l \in \mathbb{R}^s$, $l \in \mathbb{Z}$;
\item $W_1,\ldots, W_{n+1}$ are $p \times s$ weight matrices; 
\item $b_1,\ldots, b_{n+1} \in \mathbb{R}^p$ are bias vectors; 
\item $\alpha_{q, k}$, for $q = 2, \ldots, n$ and $k = 1, \ldots, n - 1$
are $p \times p$ matrices. So are $\gamma_2, \ldots, \gamma_{n}, \kappa_2, 
\ldots, \kappa_{n}$;
\item $\gamma_{n+1}$ is an $s \times s$ matrix and $\kappa_{n+1}$ is a 
$s \times p$ matrix.
\item $h \in \mathbb{R}$ is a constant and
\item $\sigma_1, \ldots, \sigma_{q+1}$ are activation function defined as
\[
\sigma_i\begin{pmatrix}x_1 \\ \vdots \\ x_p\end{pmatrix} = 
\begin{pmatrix}\sigma_i(x_1) \\ \vdots \\ \sigma_i(x_p)\end{pmatrix},
i = 1, \ldots, q+1.
\]
\end{enumerate}
\end{defn}
\noindent Before proceeding, a few remarks are in order.
\begin{rem}
There is probably a typo in equation (3) of section 2.1 Niu and other's
paper\cite{niu2019recurrent}. I believe that the range of the sum should
be $1$ to $q - 1$ as given here instead of $1$ to $n$ as mentioned in the
paper.
\end{rem}
\begin{rem}
$K_1, \ldots, K_n$ are all $p$-dimensional vectors. That is, they 
are all members of $\mathbb{R}^p$.
\end{rem}
\begin{rem}
We can readily confirm that the arguments to the activation functions
$\sigma_1, \ldots, \sigma_{q+1}$ are all $p$-dimensional vectors.
\end{rem}
\begin{rem}
The number of hidden layers is $n$. The RNN thus has $n + 2$ layers. 
\end{rem}

@@

\section{Stability analysis}\label{s9}
RNNs frequently have a problem of `exploding gradients'
\cite{gereon2018hands}. In the context of ODEs this is called an 
instability. Because of a correspondence between an RNN and an ODE, we 
can study the problem of exploding gradients in the former as the 
instability in the latter. In this section we will review certain notions
of stability of ODE.

Consider the initial value problem of equation \eqref{s3e1}. The general
$k$-step method to get its approximate numerical solution is defined by
the formula\cite{dahlquist1963special}
\begin{equation}\label{s9e1}
\sum_{i=0}^k \alpha_i x_{n+i} = h\sum_{i=0}^k \beta_i f_{n+i},
\end{equation}
where $\alpha_i, \beta_i$ are real constants.
\begin{defn}\label{s9d1}
The $k$-step method defined above is said to be A-stable
\cite{dahlquist1963special} if all solutions of \eqref{s9e1} tend to zero 
as $n \rightarrow \infty$ when the method is applied to the equation
\[
    \frac{dy}{dx} = qx
\]
where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
\eqref{s9d1}.
\end{defn}

In order to introduce the idea of B-stability and BN-stability, we need
the definition of contractive solutions\cite{lambert1991numerical}.
\begin{defn}\label{s9d2}
Let $y_1$ and $y_2$ be any two solutions of \eqref{s3e1} satisfying the
initial conditions $y_1(a) = \eta_1, y_2(a) = \eta_2, \eta_1 \ne \eta_2$.
Then if $||y_1(x_2) - y_2(x_2)|| \le ||y_1(x_1) - y_2(x_1)||$ for all
$a \le x_1 \le x_2 \le b$ then the solutions are said to be contractive
in $[a, b]$.
\end{defn}
Contractive solutions get closer to each other as one gets away from
the initial point.

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
hidden layers are not connected to each other. The Burrage-Butcher tensor 
is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
observation also corroborated by elsewhere\cite{kanai2017preventing}.
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}
